The role of cognateness in non-native spoken word recognition

### 2024-05-10 manuscript/manuscript.qmd
Final touches to first version

Listening to non-native speech is costlier than listening to native speech, even for highly proficient bilinguals, and especially in acoustically adverse situations [@lecumberri2010non; @takata1990english]. However, there are some sources of information which may make non-native speech comprehension easier. One of the sources of this increased difficulty stems from the possible mismatch between the phonology of the native and the non-native languages: some acoustic features embedded in the non-native speech signal do not overlap perfectly with any phonemic category in the listener's native language. For example, imagine a Spanish native, with no previous familiarity with any other language, who listens for the first time to French may encounter the word /`r knitr::asis_output("\U0070\U0254\U0281\U0074")`/ (*porte*), translation of door in French. The voiced uvular fricative consonant /`r knitr::asis_output("\U0281")`/ and the open-mid back rounded vowel /`r knitr::asis_output("\U0254")`/ do not exist in Spanish as phonemic categories. These discrepancies can nonetheless be perceived as allophonic variations the native (Spanish) phonemes /r/ and /o/ [@best2001discrimination], and the non-native speech signal may engage lexical processing mechanisms [@weber2004lexical], leading to the activation and selection, and therefore the recognition of its translation to Spanish /`r knitr::asis_output("\U02C8\U0070\U0077\U0065\U0072\U002E\U0074\U0061")`/ (*puerta*). The recognition of this French, unfamiliar word relies almost entirely on its similarity with its Spanish translation. Conversely, imagine now that the same Spanish listener hears English for the first time and encounters the word /`r knitr::asis_output("\U0064\U0254\U02D0")`/ (*door*). This time, phonological similarity is of little help: /`r knitr::asis_output("\U02C8\U0070\U0077\U0065\U0072\U002E\U0074\U0061")`/ and /`r knitr::asis_output("\U0064\U0254\U02D0")`/ share very few phonemes. It would come as a surprise if the listener was able to translate the English word successfully, if not by relying on previously learnt word-concept associations (which they lack, as they are unfamiliar to English).

Cross-language phonological similarity seems, at face value, an important source of information for unfamiliar listeners during auditory word recognition, even if translations are not phonologically identical. Although this mismatch has a noticeable toll on comprehension [@cutler2004patterns], the fact that word recognition can take place at all in such circumstances illustrates that non-native listeners–even low-proficiency ones–are rarely completely naïve to the language they are listening to. In the present study, we investigate the extent to which adult listeners rely on the phonological similarity between their native and non-native languages during word recognition, capitalising on the role of *cognates.*

>>>>

Listening to non-native speech is costlier than listening to native speech, even for highly proficient bilinguals, and especially in acoustically adverse situations [@lecumberri2010non; @takata1990english]. However, there are some sources of information which may make non-native speech comprehension easier. One of the sources of this increased difficulty stems from the possible mismatch between the phonology of the native and the non-native languages: some acoustic features embedded in the non-native speech signal do not overlap perfectly with any phonemic category in the listener's native language. For example, imagine a Spanish native, with no previous familiarity with any other language, who listens for the first time to French may encounter the word /pɔʁt/ (*porte*), translation of door in French. The voiced uvular fricative consonant /ʁ/ and the open-mid back rounded vowel /ɔ/ do not exist in Spanish as phonemic categories. These discrepancies can nonetheless be perceived as allophonic variations the native (Spanish) phonemes /ɾ/ and /o/ [@best2001discrimination], and the non-native speech signal may engage lexical processing mechanisms [@weber2004lexical], leading to the activation and selection, and therefore the recognition of its translation to Spanish /pweɾta/ (*puerta*). The recognition of this French, unfamiliar word relies almost entirely on its similarity with its Spanish translation. Conversely, imagine now that the same Spanish listener hears English for the first time and encounters the word /dɔː/ (*door*). This time, phonological similarity is of little help: /pweɾta/ and /dɔː/ share very few phonemes. It would come as a surprise if the listener was able to translate the English word successfully, if not by relying on previously learnt word-concept associations---which they lack, as they are unfamiliar to English.

Cross-language phonological similarity seems, at face value, an important source of information for unfamiliar listeners during auditory word recognition, even if translations are not phonologically identical. Although this mismatch has a noticeable toll on comprehension [@cutler2004patterns], the fact that word recognition can take place at all in such circumstances illustrates that non-native listeners–even low-proficiency ones–are rarely completely naïve to the language they are listening to. In the present study, we investigate the extent to which adult listeners rely on the phonological similarity between their native and non-native languages during word recognition, capitalising on the role of *cognates*.

### 2024-05-10 manuscript/manuscript.tex
Final touches to first version

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{Latin Modern Roman}
  \setmathfont[]{Latin Modern Math}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{tipa}
\usepackage{booktabs}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
\ifPDFTeX
\else
\babelfont{rm}[]{Latin Modern Roman}
\fi
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

All materials, data, and code used in this study are hosted in the Open
Science Framework \url{https://osf.io/9fjxm/} and a GitHub repository
\url{https://github.com/gongcastro/translation-elicitation.git}, along
with additional notes.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage[]{Times}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{Latin Modern Roman}
  \setmathfont[]{Latin Modern Math}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{tipa}
\usepackage{booktabs}
\DeclareTextFontCommand\textphonetic{\ipafont}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
\ifPDFTeX
\else
\babelfont{rm}[]{Latin Modern Roman}
\fi
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

All materials, data, and code used in this study are hosted in the Open
Science Framework
\href{https://osf.io/9fjxm/?view_only=aab7636ce1af48cf832596a7ea9101c5/}{https://osf.io/9fjxm/}
and a GitHub repository
\url{https://github.com/gongcastro/translation-elicitation.git}, along
with additional notes.
